% AI-Empowered Space-Air-Ground Integrated Networks (SAGIN)
% INFO 3180 Wireless Networks - Group Research Project
% Team: Divyanshu Ratti, Henry Manjo, Rajdeep Kaur

\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai2026} % DO NOT CHANGE THIS
\usepackage{times} % DO NOT CHANGE THIS
\usepackage{helvet} % DO NOT CHANGE THIS
\usepackage{courier} % DO NOT CHANGE THIS
\usepackage[hyphens]{url} % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm} % DO NOT CHANGE THIS
\usepackage{natbib} % DO NOT CHANGE THIS
\usepackage{caption} % DO NOT CHANGE THIS
\frenchspacing % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS

\pdfinfo{
/Title (AI-Empowered Space-Air-Ground Integrated Networks: A Survey)
/Author (Divyanshu Ratti, Henry Manjo, Rajdeep Kaur)
}

\title{AI-Empowered Space-Air-Ground Integrated Networks: A Survey}
\author{
    Divyanshu Ratti \and Henry Manjo \and Rajdeep Kaur \\
    Department of Computer Science / Engineering \\
    INFO 3180 -- Wireless Networks
}

\begin{document}

\maketitle

%%% -------------------------------------------------------
%%% SECTION: Abstract
%%% AUTHOR : Henry Manjo, Divyanshu Ratti
%%% -------------------------------------------------------
\begin{abstract}
Space-Air-Ground Integrated Networks (SAGIN), which combine satellite, aerial, and terrestrial communication layers, are considered a key architecture for future 6G wireless systems because they provide wider coverage and improved connectivity for applications such as disaster recovery, intelligent transportation, and Internet of Things (IoT) services. However, the mobility, heterogeneity, and dynamic nature of SAGIN make these networks difficult to manage using conventional approaches, leading to increasing interest in Artificial Intelligence (AI) techniques such as deep reinforcement learning, federated learning, agentic AI, and edge AI. This paper surveys recent research on AI-enabled SAGIN and reviews how these techniques have been applied to resource allocation, beamforming, trajectory planning, authentication, and network security. Our analysis identifies several common trends across the literature. Most AI-based solutions are validated through simulations rather than real-world deployments, researchers are increasingly adopting decentralized and autonomous network management approaches, and the majority of studies focus on performance optimization while comparatively less attention is given to security. Based on these findings, we discuss current research gaps and identify directions for advancing practical, scalable, and secure AI-enabled SAGIN for future 6G networks.
\end{abstract}

%%% -------------------------------------------------------
%%% SECTION: Introduction
%%% AUTHOR : Rajdeep Kaur
%%% -------------------------------------------------------
\section{Introduction}
% TODO (~0.5 page): Motivation for SAGIN + AI, scope of this paper,
% and a roadmap of how the paper is organized (one sentence per section).
Placeholder introduction text.

%%% END SECTION: Introduction

%%% -------------------------------------------------------
%%% SECTION: Background
%%% AUTHOR : Henry Manjo, Divyanshu Ratti, Rajdeep Kaur
%%% -------------------------------------------------------
\section{Background}
% TODO (~0.75 page): Define SAGIN (space/air/ground layers), define
% the relevant AI techniques used across the literature you reviewed
% (e.g., reinforcement learning, federated learning, agentic AI),
% and any key terminology a reader needs before the Literature Review.
Space-Air-Ground Integrated Networks (SAGIN) are communication networks that combine three different layers: space, air, and ground. The space layer includes satellites, the air layer includes platforms such as drones (UAVs) and high-altitude platforms (HAPS), and the ground layer includes cellular base stations, vehicles, IoT devices, and other terrestrial communication infrastructure. By connecting these three layers, SAGIN can provide wider coverage, improve connectivity, and support communication in areas where traditional ground networks may have limited coverage. This integrated architecture is considered an important technology for future 6G networks because it can support applications such as remote sensing, disaster recovery, intelligent transportation, and Internet of Things (IoT) services.

Throughout the literature reviewed in this survey, several recurring AI techniques are used to manage and optimize SAGIN. This section briefly defines each before they are discussed in more detail in the Literature Review.

\textbf{Deep Reinforcement Learning (DRL):} Deep reinforcement learning is an AI technique that combines reinforcement learning with deep neural networks to solve complex decision-making problems. An AI agent learns by interacting with its environment and improving its decisions based on rewards and penalties. In the papers reviewed, DRL was mainly used to optimize UAV trajectory planning, resource allocation, and task offloading while reducing energy consumption and communication delays.

\textbf{Federated Learning (FL):} Federated learning is a machine learning approach where multiple devices train the same AI model without sharing their raw data with a central server. Instead, they only exchange model updates, which helps improve privacy and reduce communication costs. Some of the papers also proposed decentralized federated learning, where satellites communicate directly with each other instead of relying on a central ground station.

\textbf{Agentic AI:} Agentic AI refers to AI systems that can make decisions and perform tasks with minimal human intervention. Instead of only following predefined instructions, these AI agents can monitor network conditions, make decisions, and adapt to changing environments automatically. In future SAGIN and 6G networks, Agentic AI can help improve resource management, routing, and overall network performance.

\textbf{Edge AI:} Edge AI is the use of AI algorithms on devices located close to where data is generated, such as UAVs, satellites, or edge servers, instead of sending all data to a central cloud. This reduces communication delays, lowers bandwidth usage, and allows faster decision-making, which is important for real-time applications in SAGIN.

%%% END SECTION: Background

%%% -------------------------------------------------------
%%% SECTION: Literature Review
%%% AUTHOR : Divyanshu Ratti
%%% -------------------------------------------------------
\section{Literature Review}
The papers reviewed in this survey address several important challenges in AI-enabled Space-Air-Ground Integrated Networks (SAGIN). Instead of focusing on one problem, researchers have explored different AI techniques to improve network performance, communication efficiency, resource management, and security. Based on the reviewed literature, the studies can be grouped into several main themes, including resource allocation and task offloading, beamforming and signal optimization, federated and decentralized learning, network security and authentication, network orchestration and handover, and general survey papers that provide an overview of the field.

\subsection{Resource Allocation and Task Offloading}
Resource allocation and task offloading are among the most widely studied topics in AI-enabled SAGIN because they directly affect communication efficiency, energy consumption, and service quality. \citet{huang2024joint} proposed a deep reinforcement learning approach to jointly optimize task offloading and resource allocation across satellites, UAVs, and cloud servers. Their method considers task dependencies and makes intelligent decisions about where computing tasks should be processed to reduce delays and improve overall network performance.

Similarly, \citet{zhang2025energy} focused on multi-UAV mobile edge computing networks, where deep reinforcement learning is used to optimize both UAV flight trajectories and computing resource allocation. Instead of solving these problems separately, the proposed method jointly optimizes them to improve energy efficiency while serving more users with lower communication delays.

\citet{jia2025dynamic} addressed a related problem by introducing a hierarchical UAV swarm architecture consisting of head UAVs and tail UAVs. Rather than using a machine learning approach, the authors combined Voronoi-based area partitioning with the Whale Optimization Algorithm to optimize UAV placement, trajectory planning, and transmit power. Their work demonstrates that optimization techniques other than machine learning can also improve the performance of aerial communication networks by balancing energy consumption and communication delay.

\subsection{Beamforming and Signal Optimization}
Another important research area in AI-enabled SAGIN is beamforming and signal optimization. Unlike the resource allocation studies, which focus on computing and communication resources, this work focuses on improving wireless signal transmission between different network nodes. \citet{khoshkbari2025distributed} proposed a distributed beamforming approach using multi-agent reinforcement learning, where multiple airborne platforms learn to coordinate their beamforming decisions without continuously exchanging channel state information. This reduces communication overhead while maintaining effective signal transmission across the network. The study demonstrates how decentralized AI techniques can improve beamforming performance and make large-scale SAGIN systems more scalable. It also highlights a growing trend in the literature toward distributed decision-making, where intelligent network nodes cooperate locally instead of relying on centralized control.

\subsection{Federated and Decentralized Learning}
Another important direction in AI-enabled SAGIN research is the use of federated and decentralized learning to improve collaboration between network nodes. Traditional federated learning often depends on a central server to collect and update AI models, but this can become inefficient in satellite networks because communication with ground stations is limited by coverage and connection time. \citet{yang2024dfedsat} addressed this challenge by proposing DFedSat, a decentralized federated learning framework that allows LEO satellites to communicate directly with each other through inter-satellite links instead of relying on a central ground station. This approach reduces communication overhead while maintaining strong learning performance and improving the robustness of the network when communication links change or some satellites become unavailable. Similar to the beamforming approach proposed by \citet{khoshkbari2025distributed}, this work reflects a growing trend toward distributed decision-making in SAGIN, where intelligent network nodes cooperate locally instead of depending on centralized control. Together, these studies suggest that decentralized AI techniques can improve the scalability, efficiency, and resilience of future SAGIN systems.

\subsection{Network Security and Authentication}
Network security and authentication have also become important research topics in AI-enabled SAGIN, although fewer studies focus on these areas compared to resource optimization and communication performance. \citet{chou2024edge} explored the use of Edge AI to improve physical layer security by allowing AI models to run closer to where data is generated, reducing communication delays while helping detect and mitigate potential security threats more quickly. Their work demonstrates how AI can strengthen wireless security without relying entirely on centralized cloud computing.

Similarly, \citet{byang2023atmas} proposed an AI-oriented two-phase multi-factor authentication framework designed specifically for SAGIN. In their approach, satellites and the network control center work together to continuously authenticate users using spatial-temporal features such as service and geographic information instead of relying only on traditional cryptographic authentication. This continuous authentication process makes the network more adaptive to the highly dynamic nature of SAGIN while improving overall security.

Together, these studies show that AI has applications beyond improving network performance. AI can also enhance authentication, privacy, and physical layer security in integrated communication networks. However, compared with the large number of studies on resource allocation, trajectory optimization, and communication efficiency, relatively fewer papers focus primarily on AI-driven security. This suggests that security remains an important research gap and an area where future AI-enabled SAGIN research could receive greater attention.

\subsection{Network Orchestration and Handover}
As SAGIN becomes more dynamic, researchers are also focusing on improving network orchestration and handover management. \citet{dev2025advanced} proposed an Agentic AI framework that uses intelligent AI agents to monitor network conditions, make autonomous decisions, and coordinate network resources with minimal human intervention. Instead of relying on fixed management rules, the proposed framework allows the network to adapt automatically to changing traffic demands and communication conditions, making it more flexible for future 6G environments.

\citet{wu2025phandover} addressed another important challenge by improving handover performance in LEO satellite networks. Because LEO satellites move quickly, user devices must frequently switch from one satellite to another, which can introduce communication delays. The authors proposed PHandover, a machine learning-assisted handover framework that predicts signal strength in advance and performs uplink and downlink handovers in parallel. Unlike many of the other studies reviewed in this survey, this work was evaluated using a real prototype with open-source 5G software and real satellite movement data, demonstrating significant improvements in handover latency. Together, these studies show that AI can improve not only resource management but also the overall coordination and reliability of highly dynamic SAGIN environments.

\subsection{Survey and Overview of AI in SAGIN}
In addition to research that proposes new algorithms and frameworks, survey papers provide a broader understanding of how AI is being applied across SAGIN. \citet{bakambekova2024interplay} reviewed a wide range of AI techniques, including machine learning, deep learning, and reinforcement learning, and examined their applications in satellites, high-altitude platforms, and UAVs. The survey covers topics such as beam management, traffic prediction, resource allocation, and network security while also discussing how SAGIN can support AI through distributed computing and data collection.

Unlike the other papers reviewed in this survey, which focus on solving specific technical problems, \citet{bakambekova2024interplay} provide an overall view of the research landscape by identifying common challenges and future research directions. The authors highlight issues such as security, scalability, and the integration of emerging 6G technologies, which are also recurring themes across the papers reviewed in this survey. As a result, this survey serves as a useful foundation for understanding the current state of AI-enabled SAGIN research and the challenges that remain before these technologies can be widely deployed.
%%% END SECTION: Literature Review

%%% -------------------------------------------------------
%%% SECTION: Discussion
%%% AUTHOR : Henry Manjo, Divyanshu Ratti, Rajdeep Kaur
%%% -------------------------------------------------------
\section{Discussion}
% TODO (~1 page): Synthesize across the literature reviewed above.
% What patterns or recurring gaps emerged (e.g., simulation-only
% evaluation, scalability, lack of real SAGIN testbeds)? Compare
% approaches against each other, not just summarize them again.
The literature reviewed in this survey reveals several recurring patterns and open challenges in applying AI to Space-Air-Ground Integrated Networks (SAGIN).

One pattern noticed across several of the papers is that most of the proposed AI solutions were only evaluated using simulations instead of real-world deployments. Although the simulation results showed improvements in areas such as resource allocation, energy efficiency, communication delay, and network performance, it is still unclear how well these methods would work in real SAGIN environments. Real networks have additional challenges such as satellite mobility, changing communication links, weather conditions, interference, and limited computing resources that are difficult to capture in simulations. This limitation appears in several papers, including the studies on multi-UAV mobile edge computing \citep{zhang2025energy}, decentralized federated learning for LEO satellite constellations \citep{yang2024dfedsat}, and deep reinforcement learning for task offloading and resource allocation \citep{huang2024joint}. This suggests that while AI-based approaches show a lot of promise, more real-world testing is needed before they can be widely deployed in future SAGIN systems.

Another pattern noticed is that many of the papers are moving away from centralized network management and instead using more decentralized or autonomous approaches. For example, the DFedSat paper removes the need for a central ground station by allowing satellites to train AI models directly with each other \citep{yang2024dfedsat}. The distributed beamforming paper uses multi-agent reinforcement learning so that each airborne platform can make decisions without constantly sharing channel state information \citep{khoshkbari2025distributed}. The Agentic AI paper also follows this trend by proposing intelligent AI agents that can monitor network conditions and make decisions automatically instead of relying on human operators \citep{dev2025advanced}. Even though these approaches use different AI techniques, they all aim to reduce communication overhead, improve scalability, and make the network more adaptable. However, they also introduce new challenges, such as coordinating many independent agents and ensuring that decentralized systems remain reliable as the network grows.

A third pattern noticed is that most of the papers focus on improving network performance rather than security. Many of the studies use AI techniques such as deep reinforcement learning, multi-agent reinforcement learning, or Agentic AI to optimize resource allocation, beamforming, task offloading, and communication efficiency. In comparison, only a few papers, such as Chou's work on Edge AI-enabled physical layer security \citep{chou2024edge}, focus mainly on protecting the network from security threats. The survey paper by Bakambekova et al. also points out that security remains one of the major open challenges in AI-enabled SAGIN \citep{bakambekova2024interplay}. This suggests that although researchers are making good progress in improving network performance, more work is still needed to develop AI solutions that can provide both high performance and strong security at the same time.

%%% END SECTION: Discussion

%%% -------------------------------------------------------
%%% SECTION: Proposed Idea
%%% AUTHOR : Divyanshu Ratti
%%% -------------------------------------------------------
\section{Proposed Idea}
% TODO (~0.75 page): Propose at least one innovative direction
% to address a gap identified in the Discussion. Does not need to be
% implemented, but should be motivated and feasible.
Placeholder proposed idea text.

%%% END SECTION: Proposed Idea

%%% -------------------------------------------------------
%%% SECTION: Conclusion
%%% AUTHOR : Divyanshu Ratti, Rajdeep Kaur
%%% -------------------------------------------------------
\section{Conclusion}
% TODO (~0.25 page): Summarize the survey and restate the proposed
% direction / future work in 3-4 sentences.
Placeholder conclusion text.

%%% END SECTION: Conclusion

%%% -------------------------------------------------------
%%% SECTION: References
%%% AUTHOR : Henry Manjo, Divyanshu Ratti, Rajdeep Kaur
%%% -------------------------------------------------------
% References are pulled from references.bib using natbib.
% Add entries to references.bib, then cite them above with \citep{key}.
\bibliography{references}
\bibliographystyle{aaai2026}

%%% END SECTION: References

\end{document}
